Generalized Rybicki Press algorithm

### 2014-10-10 paper/ESS.tex
some changes

\begin{keywords}
\end{keywords}

@@

\begin{itemize}
\item
A new $\mathcal{O}(N)$ direct solver for semi-separable matrices is obtained by embedding the semi-separable matrix into a larger banded matrix.
\item
The determinant of these semi-separable matrix is shown to equal to the determinant of the larger banded matrix, thereby enabling computing determinants of these semi-separable matrices at a computational cost of $\mathcal{O}(N)$. This is the first algorithm for computing the determinants for a general semi-separable matrix.
\item
A numerically stable generalized Rybicki Press algorithm is derived using these ideas. To be specific, fast direct solver and stable, fast determinant computation (both scaling as $\mathcal{O}(n)$) for covariance matrix of the form:
\begin{align}
A_{ij} = \sum_{l=1}^p \alpha_l \exp\left(-\beta_l \vert t_i - t_j \vert\right)
\label{eqn_covariance_matrix}
\end{align}
where $i,j \in \{1,2,\ldots,n\}$, the points $t_i$ are distinct and are distributed on an interval. The covariance matrix in Equation~\eqref{eqn_covariance_matrix} is frequently encountered in computational statistics in the context of Continuous time AutoRegressive-Moving-Average (abbreviated as CARMA) models~\cite{brockwell2002introduction, brockwell2001levy, brockwell1994continuous}.
\item
Another advantage of this algorithm from a practical view-point is that the algorithm relies only on sparse linear algebra and thereby can easily use the existing mature sparse linear algebra libraries.
\end{itemize}

The algorithm proposed in this article relies on two key ideas: (i) The covariance matrix in Equation~\eqref{eqn_covariance_matrix} is a semi-separable matrix with semi-separable rank $p$; (ii) Fast algorithms for solving and computing determinants for such semi-separable matrices can be obtained by embedding the dense covariance matrix in a large sparse banded matrix. The article also discusses how to perform the entire procedure in a numerically stable fashion. The implementation is made available at \url{https://github.com/sivaramambikasaran/ESS}~\cite{ambikasaran2014ESS} under the license provided by New York University.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%												%
%		SECTION: GENERAL SEMI-SEPARABLE			%
%												%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sparse embedding of a general semi-separable matrix}
Let $A$ be a $N \times N$ matrix, whose semi-separable rank is $p$, i.e., we have
\begin{align}
A(i,j) & =
\begin{cases}
a_{ii} & \text{ if }i=j\\
\displaystyle\sum_{l=1}^p u_j^{(l)} v_i^{(l)} & \text{ if }i>j\\
\displaystyle\sum_{l=1}^p u_i^{(l)} v_j^{(l)} & \text{ if }i<j\\
\end{cases}
\end{align}
where $i,j \in \{1,2,\ldots,N\}$. We then add the following variables $r_2^{(p)},r_3^{(p)},\ldots,r_N^{(p)}$ and $l_1^{(p)},l_2^{(p)},\ldots,l_{N-1}^{(p)}$ as before. However, unlike earlier and not surprisingly, these new variables $r_i^{(p)}$'s and $l_j^{(p)}$'s will be vectors of length $p$. Let $U_k^{(p)} = \begin{bmatrix} u_k^{(1)} & u_k^{(2)} & u_k^{(3)} & \cdots & u_k^{(p)}\end{bmatrix}$ and $V_k^{(p)} = \begin{bmatrix} v_k^{(1)} & v_k^{(2)} & v_k^{(3)} & \cdots & v_k^{(p)}\end{bmatrix}$. We then have the following relations for the additional vector variables.
\begin{align}
r_N^{(p)} = V_N^Tx_N\\
l_1^{(p)} = U_1^Tx_1
\end{align} and
\begin{align}
r_k^{(p)} & = V_k^T x_k + r_{k+1}^{(p)}\\
l_k^{(p)} & = U_k^T x_k + l_{k-1}^{(p)}
\end{align}
where $k \in \{2,\ldots,N-1\}$. Hence, we now have $(2p+1)N-2p$ variables (this includes the $N$ $x_k$'s, $N-1$ vector variables $r_k^{(p)}$ and $l_k^{(p)}$ of length $p$) and $(2p+1)N-2p$ equations relating them. Therefore, we end up with a $((2p+1)N-2p) \times ((2p+1)N-2p)$ extended sparse matrix, whose bandwidth is $(2p+1)$. This is illustrated in Figure~\ref{fig_rankm_semiseparable} for $10 \times 10$ semi-separable matrix, whose semi-separable rank is $4$.
\begin{figure}[!htbp]
\begin{center}
\includegraphics[scale=0.225]{./images/mtermsemiseparable_matrix.pdf}
\end{center}
\begin{center}
\includegraphics[scale=1]{./images/colorcode.pdf}
\end{center}
\caption{Pictorial description of the extended sparse matrix where $N=10$ and $p=4$. The color code is as follows: }
\label{fig_rankm_semiseparable}
\end{figure}

>>>>

\begin{keywords}
Semi-separable matrices, Rybicki Press algorithm, fast direct solver, fast determinant computation
\end{keywords}

@@

\begin{itemize}
\item
A new $\mathcal{O}(N)$ direct solver for semi-separable matrices is obtained by embedding the semi-separable matrix into a larger banded matrix.
\item
The determinant of these semi-separable matrix is shown to equal to the determinant of the larger banded matrix, thereby enabling computing determinants of these semi-separable matrices at a computational cost of $\mathcal{O}(N)$. This is the first algorithm for computing the determinants for a general semi-separable matrix.
\item
A numerically stable generalized Rybicki Press algorithm is derived using these ideas. To be specific, fast, stable, direct algorithms are derived for solving and computing determinants (both scaling as $\mathcal{O}(n)$) for covariance matrices of the form:
\begin{align}
A_{ij} = \sum_{l=1}^p \alpha_l \exp\left(-\beta_l \vert t_i - t_j \vert\right)
\label{eqn_covariance_matrix}
\end{align}
where $i,j \in \{1,2,\ldots,n\}$, the points $t_i$ are distinct and are distributed on an interval. The covariance matrix in Equation~\eqref{eqn_covariance_matrix} is frequently encountered in computational statistics in the context of Continuous time AutoRegressive-Moving-Average (abbreviated as CARMA) models~\cite{brockwell2002introduction, brockwell2001levy, brockwell1994continuous}.
\item
Another advantage of this algorithm from a practical view-point is that the algorithm relies only on sparse linear algebra and thereby can easily use the existing mature sparse linear algebra libraries.
\end{itemize}

The algorithm discussed in this article has been implemented in C++ and the implementation is made available at \url{https://github.com/sivaramambikasaran/ESS}~\cite{ambikasaran2014ESS} under the license provided by New York University.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%												%
%		SECTION: GENERAL SEMI-SEPARABLE			%
%												%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sparse embedding of a general semi-separable matrix}
Let $A$ be a $N \times N$ matrix, whose semi-separable rank is $p$, i.e., we have
\begin{align}
A(i,j) & =
\begin{cases}
a_{ii} & \text{ if }i=j\\
\displaystyle\sum_{l=1}^p u_j^{(l)} v_i^{(l)} & \text{ if }i>j\\
\displaystyle\sum_{l=1}^p u_i^{(l)} v_j^{(l)} & \text{ if }i<j\\
\end{cases}
\end{align}
where $i,j \in \{1,2,\ldots,N\}$. We then add the following variables $r_2^{(p)},r_3^{(p)},\ldots,r_N^{(p)}$ and $l_1^{(p)},l_2^{(p)},\ldots,l_{N-1}^{(p)}$ as before. However, not surprisingly, these new variables $r_i^{(p)}$'s and $l_j^{(p)}$'s will be vectors of length $p$. Let $U_k^{(p)} = \begin{bmatrix} u_k^{(1)} & u_k^{(2)} & u_k^{(3)} & \cdots & u_k^{(p)}\end{bmatrix}$ and $V_k^{(p)} = \begin{bmatrix} v_k^{(1)} & v_k^{(2)} & v_k^{(3)} & \cdots & v_k^{(p)}\end{bmatrix}$. We then have the following relations for the additional vector variables.
\begin{align}
r_N^{(p)} = V_N^Tx_N\\
l_1^{(p)} = U_1^Tx_1
\end{align} and
\begin{align}
r_k^{(p)} & = V_k^T x_k + r_{k+1}^{(p)}\\
l_k^{(p)} & = U_k^T x_k + l_{k-1}^{(p)}
\end{align}
where $k \in \{2,\ldots,N-1\}$. Hence, we now have $(2p+1)N-2p$ variables (this includes the $N$ $x_k$'s, $N-1$ vector variables $r_k^{(p)}$ and $l_k^{(p)}$ of length $p$) and $(2p+1)N-2p$ equations relating them. Therefore, we end up with a $((2p+1)N-2p) \times ((2p+1)N-2p)$ extended sparse matrix, whose bandwidth is $(2p+1)$. This is illustrated in Figure~\ref{fig_rankm_semiseparable} for $10 \times 10$ semi-separable matrix, whose semi-separable rank is $4$.
\begin{figure}[!htbp]
\begin{center}
\includegraphics[scale=0.225]{./images/mtermsemiseparable_matrix.pdf}
\end{center}
\begin{center}
\includegraphics[scale=1]{./images/colorcode.pdf}
\end{center}
\caption{Pictorial description of the extended sparse matrix where $N=10$ and $p=4$. The color code is as follows: }
\label{fig_rankm_semiseparable}
\end{figure}
